---
title: "Convert single PLINK 2 output gwas files from HG19 to HG38 using liftOver"
output: html_notebook
---

## Project: GP2_EUR_YOPD_metaGWAS

### Author: 
- MV, adapted by IHL

### Version: 
- R version 4.4.1 (2024-06-14)

### Notebook Overview
- This notebook shows how the single PLINK 2 output gwas files were lifted from HG19 to HG38 using liftOver
- There are no indels in the PLINK2 output gwas files

### Changelog
- 23.10.25 Notebook made 

### Read in the necessary libraries
```{r}
library(dplyr)
library(GenomicRanges)
library(rtracklayer)
library(Biostrings)
library(BSgenome.Hsapiens.UCSC.hg38)
library(data.table)

```

### Read in the input file to be lifted over
```{r}
#### 0. INPUT  -----------------------------------------

# The IPDGC SPAIN GWAS
df <- fread(
  "~/Documents/GWAS_recessive_SPAIN.PHENO1.glm.logistic.hybrid", 
  sep = "\t", 
  colClasses = c("ID" = "character"))  # Specify ID column as character



head(df)
dim(df)
```

### Rename columns to fit with the liftOver script
```{r}

# Required columns: id, chrom (without "chr"), pos, ref, alt


names(df)[names(df) == "#CHROM"] <- "chrom"
names(df)[names(df) == "ID"] <- "id"
names(df)[names(df) == "POS"] <- "pos"
names(df)[names(df) == "REF"] <- "ref"
names(df)[names(df) == "ALT"] <- "alt"

```

## Perform LiftOver
```{r}

#### 1. MINIMAL VARIANT TABLE --------------------------
mini <- df |> select(id, chrom, pos, ref, alt)

gr37 <- GRanges(
  seqnames = paste0("chr", mini$chrom),
  ranges   = IRanges(start = mini$pos,
                     width = pmax(nchar(mini$ref), 1L)),  # robust for indels
  strand   = "*",
  id  = mini$id,
  ref = mini$ref,
  alt = mini$alt
)

```


```{r}

#### 2. LIFTOVER ---------------------------------------
chain <- import.chain("/Users/Documents/hg19ToHg38.over.chain")

lo <- liftOver(gr37, chain)

# keep variants that map *exactly once*
mapped_ok   <- elementNROWS(lo) == 1L
gr38        <- unlist(lo[mapped_ok])

```

```{r}
#### 3. ALLELE FIX-UP ----------------------------------
# Fetch hg38 reference base(s) at lifted positions
ref38_base <- as.character(
  getSeq(BSgenome.Hsapiens.UCSC.hg38,
         names  = seqnames(gr38),
         start  = start(gr38),
         width  = nchar(mcols(gr38)$ref))
)

# Helper: reverse-complement any allele(s)
revcomp <- function(x) as.character(reverseComplement(DNAStringSet(x)))

r   <- mcols(gr38)$ref           # original REF
a   <- mcols(gr38)$alt           # original ALT
ref <- ref38_base                # hg38 reference

# Identify how to reconcile REF with hg38
is_match      <-  r == ref
is_swap       <-  a == ref
is_flip       <-  revcomp(r) == ref
is_flip_swap  <-  revcomp(a) == ref

actions <- ifelse(is_match,       "none",
           ifelse(is_swap,        "swap",
           ifelse(is_flip_swap,   "flip+swap",
           ifelse(is_flip,        "flip", "ref_mismatch"))))

# Generate final hg38-compatible alleles
new_alt <- dplyr::case_when(
  actions == "none"       ~ a,
  actions == "swap"       ~ r,
  actions == "flip"       ~ revcomp(a),
  actions == "flip+swap"  ~ revcomp(r),
  TRUE                    ~ NA_character_        # ref_mismatch or other junk
)

new_ref <- ref38_base




```


```{r}
#### 4. COLLECT RESULTS & AUDIT -------------------------
# variants that passed the checks
keep <- actions != "ref_mismatch"
lifted <- tibble(
  id      = mcols(gr38)$id[keep],
  chrom38 = as.character(sub("^chr", "", seqnames(gr38)[keep])),
  pos38   = start(gr38)[keep],
  ref38   = new_ref[keep],
  alt38   = new_alt[keep],
  action  = actions[keep]
)

# variants that failed at any stage
dropped <- tibble(
  id     = c(mcols(gr37)$id[!mapped_ok],
             mcols(gr38)$id[ actions == "ref_mismatch" ]),
  reason = c(rep("unmapped_or_multimapped", sum(!mapped_ok)),
             rep("ref_mismatch", sum(actions == "ref_mismatch")))
)


```


```{r}
#### 5. MERGE BACK WITH ORIGINAL DATA -------------------
df38 <- df |>               # original big data frame
  left_join(lifted, by = "id")


```


```{r}
#### 6. OPTIONAL: WRITE LOG FILES -----------------------
#write.table(lifted,  "lifted_variants.tsv",  sep = "\t", quote = FALSE, row.names = FALSE)
#write.table(dropped, "dropped_variants.tsv", sep = "\t", quote = FALSE, row.names = FALSE)

```

```{r}
#### 7. UPDATE THE "A1" AND "OMITTED" COLUMNS
A1_38 <- ifelse(df38$action == "flip" | df38$action == "flip+swap", revcomp(df38$A1), df38$A1)
OMITTED_38 <- ifelse(df38$action == "flip" | df38$action == "flip+swap", revcomp(df38$OMITTED), df38$OMITTED) 
df38$A1_38 <- A1_38
df38$OMITTED_38 <- OMITTED_38
```


### Inspect the swapped, flipped and flipped+swapped alleles after liftover 
```{r}
print("look at the swapped")
look_swap <- df38 %>% filter(action == "swap")
head(look_swap)
```
```{r}
print("look at the flipped")
look_flip <- df38 %>% filter(action == "flip")
head(look_flip)
```
```{r}

print("look at the flip+swapped")
look_flipswap <- df38 %>% filter(action == "flip+swap")
head(look_flipswap)

```

### Look at the "dropped" IDS
```{r}
print("look at the dropped")
head(dropped)
table(dropped$reason)
```

### Inspect at the final dataframe
```{r}
colnames(df38)
dim(df38)
```

## Write the df38 including the hg19 original coordinates and alleles and the lifted hg38 coordinates and alleles to a .tsv file using fwrite 
```{r}
# Specify the path where you want to save your .tsv file
output_path <- "/Users/Documents/lifted_from_hg19_to_hg38_GWAS_recessive_SPAIN.PHENO1.glm.logistic.hybrid.tsv"

# Write the data.table to a .tsv file
fwrite(df38, file = output_path, sep = "\t", na = "NA")

# Print a message to confirm saving
cat("File saved to:", output_path, "\n")
```

```{r}
#Optional
# Specify the path where you want to save your .tsv file
#output_path <- "/Users/Documents/DROPPED_from_hg19_to_hg38_GWAS_recessive_SPAIN.PHENO1.glm.logistic.hybrid.tsv"

# Write the data.table to a .tsv file
#fwrite(dropped, file = output_path, sep = "\t", na = "NA")

# Print a message to confirm saving
#cat("File saved to:", output_path, "\n")
```
