---
title: "Remove extreme outliers from recessive single GWAS files prior to running METAL"
output: html_notebook
---

## Project: GP2_EUR_YOPD_metaGWAS

### Author: 
- IHL

### Version: 
- R version 4.5.2 (2025-10-31)

### Notebook Overview
- METAL is not able to run properly with extreme outlier values, including infinite values
- This script displays how SNPs with extreme outlier values (beta-values >4 and <-4) were removed from the single, recessive GWAS analyses prior to METAL meta-analysis.
- The same steps were conducted on all single, recessive GWAS files formatted for METAL

### Changelog
- 05.11.25 Notebook made 

##Load R-packages
```{r}
library(tidyverse)
library(data.table)
```

## Read in the file
```{r}

gp2 <- fread("/Users/Documents/GP2_recessive_gwas_for_METAL.tsv")

head(gp2)
dim(gp2)
```
# Inspect the rows with beta-values >4, similar to OR ~55 and <-4, similar to OR ~ 0.02
```{r}
# Assuming your dataframes are stored in a list
cohorts <- list(gp2)

# Assign cohort names to the list for easier access
cohort_names <- c("gp2")
names(cohorts) <- cohort_names

# Loop through each cohort and create a new dataframe with both large and small BETA values
for (cohort_name in cohort_names) {
  
  # Ensure we are dealing with a dataframe
  if (is.data.frame(cohorts[[cohort_name]])) {
    
    # Get the original dataframe
    original_df <- cohorts[[cohort_name]]
    
    # Filter for high BETA values
    high_beta_df <- original_df[original_df$BETA > 4, ]
    
    # Filter for low BETA values
    low_beta_df <- original_df[original_df$BETA < -4, ]
    
    # Add a column to identify large and small categories
    high_beta_df$Beta_Type <- "large"
    low_beta_df$Beta_Type <- "small"
    
    # Combine the two dataframes into one
    combined_df <- rbind(high_beta_df, low_beta_df)
    
    # Create a new name for the combined dataframe
    combined_df_name <- paste0("combined_B_", cohort_name)
    
    # Assign the combined dataframe to a new variable
    assign(combined_df_name, combined_df)
    
  } else {
    warning(paste(cohort_name, "is not a dataframe."))
  }
}

```

#Inspect the dimensions of the files with large or small values
```{r}
dim(combined_B_gp2)
head(combined_B_gp2)

```

# Create new .tsv files without the SNPs with outlier BETA-values
```{r}
# Assuming your dataframes are stored in a list
cohorts <- list(gp2)

# Assign cohort names to the list for easier access
cohort_names <- c("gp2")
names(cohorts) <- cohort_names

# Specify the directory where you want to save the new dataframes
output_directory <- "/Users/Documents/Results/"  # Modify this path as needed

# Loop through each cohort and create new dataframes with the filtered data
for (cohort_name in cohort_names) {
  
  # Ensure we are dealing with a dataframe
  if (is.data.frame(cohorts[[cohort_name]])) {
    
    # Get the original dataframe
    original_df <- cohorts[[cohort_name]]
    
    # Filter the dataframe to exclude BETA values above 4 and below -4
    filtered_df <- original_df[original_df$BETA >= -4 & original_df$BETA <= 4, ]
    
    # Create a new name for the filtered dataframe
    new_df_name <- paste0(cohort_name, "_between_minus4_and_4_recessive_gwas.tsv")
    
    # Construct the full file path
    file_path <- file.path(output_directory, new_df_name)
    
    # Save the filtered dataframe to a TSV file without quotes around the header names
    write.table(filtered_df, file_path, sep = "\t", row.names = FALSE, quote = FALSE)
    
  } else {
    warning(paste(cohort_name, "is not a dataframe."))
  }
}

# The filtered dataframes will be saved in the specified directory

dim(filtered_df)
```