---
title: "Investigate hits in the additive YOPD meta-GWAS analysis"
output:
  html_document:
    df_print: paged
---

## Project: GP2_EUR_YOPD_metaGWAS

### Author: 
- IHL

### Version: 
- R version 4.4.1 (2024-06-14)

### Notebook Overview
- This script displays genome wide significant SNPs in the additive meta-analysis. 
- The same method was used to display genome wide significant SNPs in the recessive meta-analysis

### Changelog
- Notebook generated 23.10.25

#Load libraries
```{r}
library(tidyverse)
library(data.table)
library(Hmisc)
library(dplyr)
library(tidyr)
library(qqman)
```

#Read in dataframes
```{r}
META <- fread('~/Documents/META_additive.tbl')

#change P-value to P for simplicity
META <- META %>%
 rename(P = `P-value`)
```

```{r}
# Separate the ID column into CHR and POS, keeping the ID column untouched
META <- META %>%
  separate(MarkerName, into = c("CHR", "POS"), sep = ":", convert = TRUE, remove = FALSE)

head(META)

```

```{r}
#Look at genome wide significant SNPs:

# Find the significant p-values
significant_p_values_orig <- META %>% filter(P < 0.00000005) %>% arrange(P)

# This will give you the rows where P values are below the threshold.
print(significant_p_values_orig)


```

#Look at the significant hits NOT on chromosome 4 
```{r}
sign_not_4 <- META %>% filter(P < 0.00000005) %>% arrange(P) 
sign_not_4 <- sign_not_4 %>% filter(!(CHR == "4"))
print(sign_not_4)
```
