---
title: "Meta-analysis Figures"
output: html_notebook
---
### Project: GP2_EUR_YOPD_metaGWAS

### Author:IHL

### Version:R version 4.5.2 (2025-10-31)

### Notebook Overview
- This script displays the code to generate the Manhattan-plot, Miami-plot, Beta-beta-plots, and Forest plots

### Changelog
- Notebook started 23.10.25

# Libraries
```{r}
library(tidyverse)
library(data.table)
library(qqman)
library(ggplot2)
```


# Manhattan plot and QQ-plot

### Read in and prepare the metaGWAS output (METAL-format)
- Make a custom dataset ("add") before making the plots for the additive meta analysis output
- The same steps were taken for the recessive METAL-meta analysis output to generate the file "recc"
```{r}
# Read in file
META <- fread(additive_metaanalysis.tbl) 

# Select the variables you need to make the plot
add <- META %>% select(CHR, POS, MarkerName, P)

#Rename files to work with qqman R package
add <- data %>%
  rename(
    BP = POS,
    SNP = MarkerName,
    )
```


### Make the Manhattan plot 
```{r}
library(qqman)


tiff(
  filename = "additive_manhattan.tiff",
  width    = 18.5,  # 185 mm
  height   = 12,    # adjust if needed
  units    = "cm",
  res      = 300
)

manhattan(add, 
          col = c("blue4", "orange3"), 
          cex = 0.6, cex.axis = 0.8, 
          las = 1)

dev.off()
```

### Make the QQ plot 
```{r}
tiff(
  filename = "additive_QQ_plot.tiff",
  width    = 18.5,  # 185 mm
  height   = 12,    # adjust if needed
  units    = "cm",
  res      = 300
)

qq(add$P)

dev.off()
```


# Miami plot (Additive on top, recessive at the bottow)
```{r}
## ====================================================================
## FULL SCRIPT: Miami plot with Locus annotations, no grid,
## and dashed genome-wide line at p < 5e-8 (red) in both panels
## ====================================================================

library(dplyr)
library(ggplot2)
library(ggrepel)
library(patchwork)

## --------------------------------------------------------------------
## 0. Your GWAS dataframes: add, recc. Your list of top hits add_hits, recc_hits 
## --------------------------------------------------------------------
## These must already exist in your R session.
## add and recc must contain at least: CHR, BP, P
## add_hits and recc_hits must contain CHR, BP, Locus (SNP)

# Ensure types
add  <- add %>%
  mutate(
    CHR = as.integer(CHR),
    BP  = as.numeric(BP),
    P   = as.numeric(P)
  )

recc <- recc %>%
  mutate(
    CHR = as.integer(CHR),
    BP  = as.numeric(BP),
    P   = as.numeric(P)
  )


## --------------------------------------------------------------------
## 2. Function: cumulative genomic position + -log10(P)
## --------------------------------------------------------------------

make_manh_data <- function(df) {
  df <- df %>%
    arrange(CHR, BP) %>%
    group_by(CHR) %>%
    mutate(chr_max = max(BP, na.rm = TRUE)) %>%
    ungroup()

  chr_offsets <- df %>%
    group_by(CHR) %>%
    summarise(chr_len = max(BP, na.rm = TRUE), .groups = "drop") %>%
    arrange(CHR) %>%
    mutate(offset = lag(cumsum(chr_len), default = 0)) %>%
    select(CHR, offset, chr_len)

  df <- df %>%
    left_join(chr_offsets, by = "CHR") %>%
    mutate(
      pos  = BP + offset,   # cumulative genomic position
      logp = -log10(P)      # -log10(P)
    )

  list(
    data     = df,
    chr_info = chr_offsets
  )
}

## --------------------------------------------------------------------
## 3. Apply to additive and recessive data
## --------------------------------------------------------------------

add_m   <- make_manh_data(add)
recc_m  <- make_manh_data(recc)

add_df   <- add_m$data
recc_df  <- recc_m$data
chr_info <- add_m$chr_info

## --------------------------------------------------------------------
## 4. Attach plotting coordinates to hits using CHR + BP
## --------------------------------------------------------------------

add_hits_pos <- add_hits %>%
  mutate(
    CHR = as.integer(CHR),
    BP  = as.numeric(BP)
  ) %>%
  inner_join(
    add_df %>% select(CHR, BP, pos, logp),
    by = c("CHR", "BP")
  )

recc_hits_pos <- recc_hits %>%
  inner_join(
    recc_df %>% select(CHR, BP, pos, logp),
    by = c("CHR", "BP")
  )

## --------------------------------------------------------------------
## 5. Colors, y limits, chromosome centers, genome-wide line
## --------------------------------------------------------------------

add_df <- add_df %>%
  mutate(col = ifelse(CHR %% 2 == 0, "blue4", "orange3"))

recc_df <- recc_df %>%
  mutate(col = ifelse(CHR %% 2 == 0, "blue4", "orange3"))

y_max <- max(
  max(add_df$logp,  na.rm = TRUE),
  max(recc_df$logp, na.rm = TRUE)
) + 1
y_max <- max(y_max, 15)

chr_centers <- chr_info %>%
  mutate(center = offset + chr_len / 2)

# Genome-wide significance line: p = 5e-8
gw_line <- -log10(5e-8)

## --------------------------------------------------------------------
## 6. Top panel: additive with Locus labels, no grid, dashed red GW line
## --------------------------------------------------------------------

p_top <- ggplot(add_df, aes(x = pos, y = logp)) +
  geom_point(aes(color = col), size = 0.6) +
  scale_color_identity() +
  scale_y_continuous(
    limits = c(0, y_max),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_x_continuous(
    breaks = chr_centers$center,
    labels = chr_centers$CHR
  ) +
  # genome-wide significance line (dashed)
  geom_hline(yintercept = gw_line, color = "red", linetype = "dashed") +
  ylab(expression(-log[10](P))) +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x  = element_blank(),   # remove horizontal x-axis line in top panel
    panel.grid   = element_blank(),
    panel.border = element_blank()
  ) +
  geom_text_repel(
    data = add_hits_pos,
    aes(x = pos, y = logp, label = Locus),
    size = 2.2,
    nudge_y = 0.5,
    min.segment.length = 0,
    max.overlaps = Inf,
    box.padding = 0.25
  )

## --------------------------------------------------------------------
## 7. Bottom panel: recessive (mirrored) with Locus labels, no grid, dashed red GW line
## --------------------------------------------------------------------

p_bottom <- ggplot(recc_df, aes(x = pos, y = -logp)) +
  geom_point(aes(color = col), size = 0.6) +
  scale_color_identity() +
  scale_y_continuous(
    limits = c(-y_max, 0),
    expand = expansion(mult = c(0.02, 0))
  ) +
  scale_x_continuous(
    breaks = chr_centers$center,
    labels = chr_centers$CHR
  ) +
  # mirrored genome-wide significance line (dashed, negative)
  geom_hline(yintercept = -gw_line, color = "red", linetype = "dashed") +
  xlab("Chromosome") +
  ylab(expression(-log[10](P))) +
  theme_classic() +
  theme(
    panel.grid   = element_blank(),
    panel.border = element_blank()
  ) +
  geom_text_repel(
    data = recc_hits_pos,
    aes(x = pos, y = -logp, label = Locus),
    size = 2.2,
    nudge_y = -0.5,
    min.segment.length = 0,
    max.overlaps = Inf,
    box.padding = 0.25
  )

## --------------------------------------------------------------------
## 8. Combine and save
## --------------------------------------------------------------------

miami_plot <- p_top / p_bottom + plot_layout(heights = c(1, 1))

ggsave(
  filename = "/Users//Documents/miamiplot.tiff",
  plot     = miami_plot,
  width    = 18.5,   # cm
  height   = 12,     # cm
  units    = "cm",
  dpi      = 300
)

```



# Beta-beta-plots

###  Read in files for the beta-beta-plot
- Input data frame must include SNP identifiers and the two beta (effect size) columns for each of the two different summary statistics.
- Ensure effect allele, non-effect allele, allele frequency (if used), and effect sizes are aligned between the two summary-statistic files before running.

```{r}
# Read in the dataframe
df_aligned <- fread("df_aligned.tsv")

```

### Make the beta-beta plot
```{r}

# Load ggplot2 library
library(ggplot2)


# Create the scatter plot with semi-transparent points
ggplot(df_aligned, aes(x = Effect_Nalls, y = Effect_YOPD)) + 
  geom_point(color = "blue", size = 3, alpha = 0.5) +  # Points with 50% transparency
  geom_hline(yintercept = 0, color = "red", linetype = "solid", lwd = 1) +  # Red line at y=0
  geom_vline(xintercept = 0, color = "red", linetype = "solid", lwd = 1) +  # Red line at x=0
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", lwd = 1) +  # 45-degree red line
  labs(x = "Effect (Beta), Nalls 2019)", 
       y = "Effect (Beta), YOPD)",
       title = "B)") +  # Axis labels and title
  theme_minimal() +  # Minimal theme
  ylim(-1, 1)    +   # Set limits for the y-axis from -1 to +1
xlim(-1,1)
```

# Forest Plot 
- Example: SNCA risk locus, 4:89704960 (chr:pos) 

### Read in the single GWASes used in the meta-analysis
```{r}
gp2 <- fread("gp2.tsv")
ngrc <- fread("ngrc.tsv")
dutch <- fread("dutch.tsv")
germany <- fread("germany.tsv")
neurox <- fread("neurox.tsv")
nia <- fread("nia.tsv")
oslo <- fread("oslo.tsv")
spain <-fread("spain.tsv")
tubi <- fread("tubi.tsv")
```

###  Extract the rows with the target SNP info
```{r}
# List of your data frame names
cohorts <- c("gp2", "ngrc", "dutch", "germany", "neurox", "nia", "oslo", "spain", "tubi")

# SNP ID to match
target_snp <- "4:89704960"

# Loop over each data frame
for (cohort in cohorts) {
  # Get the actual data frame
  df <- get(cohort)
  
  # Subset the row(s) where ID matches the SNP
  snp_row <- df[df$ID == target_snp, ]
  
  # Create a new variable like gp2_snp, etc.
  assign(paste0(cohort, "_snp"), snp_row)
}

snp_df <- rbind(gp2_snp, ngrc_snp, dutch_snp, germany_snp,neurox_snp, nia_snp, oslo_snp, spain_snp, tubi_snp)

print(snp_df)
```
Note - ensure that the effect allele, non-effect allele, frequency of effect allele and effect sizes are aligned for all cohorts. If not, this must be fixed prior to making the figure

### Make the figure
```{r}

labs <- snp_df$COHORT
yi <- snp_df$BETA
sei <- snp_df$SE
res  <- rma(yi=yi, sei=sei, method="FE")
summary(res)

# Forest plot
forest(
  res,
  transf = exp,                # show odds ratios instead of log-odds
  refline = 1,                 # reference line at OR = 1
  xlab = "Odds Ratio (95% CI)", 
  slab = labs,                 # study labels
  mlab = "Summary Estimate",   # summary label
  main = "rs356182 (SNCA)" # <-- Adds the header/title
)

```
